Measuring causal influence with\\ back-to-back regression: the linear case

### 2019-05-23 paper/meg_exps.tex
typos

We attempt to identify which word features ($X$) is reliably associated with the MEG recordings $Y$ at each time sample and for each subject separately. To this aim,
we compare three methods: forward regression, backward regression, and back-to-back regression.
Corresponding results can be found in Figure \ref{fig:meg_twocurves}.

@@

Forward regression was performed with an $\ell_2$-regularized least-squares regression $H \in \mathbb{R}^{d_x \times d_y}$ from $X$ to $Y$, and estimates the causal influence $E_{i,i}$ as $\| H_{i, :} \|^2$.
%
Backward regression was performed with an $\ell_2$-regularized least-squares regression $G \in \mathbb{R}^{d_y \times d_x}$ from $Y$ to $X$, and estimates the causal influence $E_{i, i}$ as the Pearson correlation between the $i$-th column of $X$ and the $i$-th column of $YG$, using a validation sample.
%
Finally, we used the back-to-back regression was performed using two $\ell_2$-regularized
least squares regressions $G$ and $H$ as described in section \ref{sec:algorithm}, using a bagging of 10 alphas logarithmically distributed between -5 and 5.

@@

The three types of analyses were implemented for each subject and each time
sample independently. To estimate the reliability of these estimates in the
population, we tested, using one-sample $t$-test whether the distribution of
the estimated contribution of each factor different from baseline ($t=0$ms),
as $\text{t-test}(\hat E_{t=0} - \hat E_{\tau})$, with $\tau \in $ ranging from
150ms to 1s after word onset. We use the resulting t and p-values as proxy to
quantify the performance of each model. These results are depicted in Figure
4 of the Supplementary Material. For concision purposes, we only report this metric averaged
between 150ms and 1s after word onset.

@@

Finally we investigate in Figure \ref{fig:megresult} whether B2B was sufficiently sensitive to reliably detect a multiplicity of subtle features, such as individual letters and part-of-speech. The vast majority the visual and lexical features appeared to be reliably detected (see Figure 3 in Supplementary Material). B2B thus promises to improve the detectability of subtle effects.

>>>>

We attempt to identify which word features ($X$) are reliably associated with the MEG recordings $Y$ at each time sample and for each subject separately. To this aim,
we compare three methods: forward regression, backward regression, and back-to-back regression.
Corresponding results can be found in Figure \ref{fig:meg_twocurves}.

@@

Forward regression was performed with an $\ell_2$-regularized least-squares regression $H \in \mathbb{R}^{d_x \times d_y}$ from $X$ to $Y$, and estimates the causal influence $E_{i,i}$ as $\| H_{i, :} \|^2$.
%
Backward regression was performed with an $\ell_2$-regularized least-squares regression $G \in \mathbb{R}^{d_y \times d_x}$ from $Y$ to $X$, and estimates the causal influence $E_{i, i}$ as the Pearson correlation between the $i$-th column of $X$ and the $i$-th column of $YG$, using a validation sample.
%
Finally, we used the back-to-back regression was performed using two $\ell_2$-regularized
least squares regressions $G$ and $H$ as described in section \ref{sec:algorithm}, using a bagging factor of 10, and alpha cross-validated between $10^{-5}$ and $10^{5}$.

@@

The three types of analyses were implemented for each subject and each time
sample independently. To estimate the reliability of these estimates in the
population, we tested, using a one-sample $t$-test whether the distribution of
the estimated contribution of each factor differed from baseline ($t=0$ms),
as $\text{t-test}(\hat E_{t=0} - \hat E_{\tau})$, with $\tau$ ranging from
150ms to 1s after word onset. We use the resulting t and p-values as proxy to
quantify the performance of each model. These results are depicted in Figure
3 of the Supplementary Material. For concision purposes, we only report this metric averaged
between 150ms and 1s after word onset.

@@

Finally we investigate in Figure \ref{fig:megresult} whether B2B was sufficiently sensitive to reliably detect a multiplicity of subtle features, such as individual letters and part-of-speech. The vast majority of the visual and lexical features appeared to be reliably detected (see Figure 4 in Supplementary Material). B2B thus promises to improve the detectability of subtle effects.

### 2019-05-23 paper/paper.tex
typos

More generally, B2B is related to CCA \citep{cca_hotelling} in that it can be thought of as a cross-decomposition of $X$ and $Y$. B2B departs from CCA in four main aspect. First, CCA performs an additional eigenvector decomposition what would be the $H$ coefficients in B2B \eqref{eq:solH}. Second, CCA does not typically optimize two distinct regularization parameters. Third CCA does not use bagging between $G$ and $H$. Finally, CCA does not diagonalize nor binarize the $H$ matrix to estimate $\hat E$. More generally, CCA aims to find maximally correlated components between $X$ and $Y$, and does not assume any causal direction. By contrast, B2B assumes that $X$ causes $Y$. Together, these differences may explain why B2B can reliably outperform CCA on estimating causal influences and predicting $Y$ in novel environments \ref{table:synthetic}.
%
% If all the associated eigenvalues are either $0$ or $1$, then all the variables are either part of the $Y$ subspace, or orthogonal to it, respectively.
% %
% This is because these eigenvalues correspond to the squared cosine of the angle between the corresponding variable in $X$ and the $Y$ subspace.
% %
% Thus, B2B is equivalent to CCA under the constraint that the subspaces spanned by $X$ and $Y$ are perpendicular, $X$ is spanned by vectors of $Y$ and vectors orthogonal to $Y$, and that the parallel and orthogonal components of $X$ are spanned by disjoint sets of features of $X$.

@@

B2B is also related, but different, from usual causal discovery algorithms \citep{peters2017elements} that hypothesize on the direction of causality, or from measures of causal influence requiring time information \citep{granger1969investigating, janzing2013quantifying}.
%
B2B seems to have a similar flavour to half-sibling regression \citep{scholkopf2016modeling}, in the sense that we observe a large number of variables corrupted by the same source of noise; we leave the study of the precise link between these two models for future work.

>>>>

More generally, B2B is related to CCA \citep{cca_hotelling} in that it can be thought of as a cross-decomposition of $X$ and $Y$. B2B departs from CCA in four main aspects. First, CCA performs an additional eigenvector decomposition what would be the $H$ coefficients in B2B \eqref{eq:solH}. Second, CCA does not typically optimize two distinct regularization parameters. Third, CCA does not use bagging between $G$ and $H$. Finally, CCA does not diagonalize nor binarize the $H$ matrix to estimate $\hat E$. More generally, CCA aims to find maximally correlated components between $X$ and $Y$, and does not assume any causal direction. By contrast, B2B assumes that $X$ causes $Y$. Together, these differences may explain why B2B can reliably outperform CCA on estimating causal influences and predicting $Y$ in novel environments \ref{table:synthetic}.
%
% If all the associated eigenvalues are either $0$ or $1$, then all the variables are either part of the $Y$ subspace, or orthogonal to it, respectively.
% %
% This is because these eigenvalues correspond to the squared cosine of the angle between the corresponding variable in $X$ and the $Y$ subspace.
% %
% Thus, B2B is equivalent to CCA under the constraint that the subspaces spanned by $X$ and $Y$ are perpendicular, $X$ is spanned by vectors of $Y$ and vectors orthogonal to $Y$, and that the parallel and orthogonal components of $X$ are spanned by disjoint sets of features of $X$.

@@

B2B is also related to, but different from usual causal discovery algorithms \citep{peters2017elements} that hypothesize on the direction of causality, or from measures of causal influence requiring time information \citep{granger1969investigating, janzing2013quantifying}.
%
B2B seems to have a similar flavour to half-sibling regression \citep{scholkopf2016modeling}, in the sense that we observe a large number of variables corrupted by the same source of noise; we leave the study of the precise link between these two models for future work.
